---
title: "README"
author: "Yue Pan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document
---

```{r setup, include=FALSE}
# library(ggplot2)
# library(dplyr)
# library(purrr)
# library(here)
# 
# source(here("R/utils.R"))
# source(here("R/qq_env_small.R"))
# source(here("R/data_representation.R"))
# source(here("R/hclustDepart.R"))
# source(here("R/seurat_plus.R"))

```


## Overview

The R package scppp is developed to visualize the Poissoneity of scRNA-seq data, and explore cell clustering based on departure as a novel data representation.

## Installation

Please install from CRAN ???.

```{r install, eval = FALSE}
#utils::install.packages("scppp")
library(scpoissonmodel)
```

## Visualization

The examples below show Q-Q envelope plots comparing sample data with a theoretical Poisson distribution given the Poisson parameter (mean). The input can be either a numeric vector or a numeric matrix consists of integers. If the input is a numeric matrix, the GLM-PCA algorithm will be applied for parameter estimation. Then some matrix entries (default 200) will be selected as sample data and compare with the theoretical distribution. 

The code here is specific for validation of independent Poisson distributions, but such idea can be applied to different types of data under different assumptions of distributions.

```{r example1, message=FALSE, warning=FALSE}
# Numeric vector as input
set.seed(1234)
scppp_obj <- scppp(rpois(200, 3L))
qqplot_env_pois(scppp_obj, 3, 100)

# Count matrix as input
set.seed(1234)
dat <- matrix(c(rpois(300, 5L), rpois(200, 1L)), ncol = 20)
scppp_obj <- scppp(dat)
qqplot_env_pois(scppp_obj, L = 2, lambda = 5) # small L for a data with less variance
  
```

Please refer to vignette for a more detailed guide, including the application in the real data.

## Clustering

The example below shows clustering pipeline based on a simplified two-cluster data. The results are save in a list. The first element contains the clustering results for each sample, where the cluster label shows the top-down cluster in a tree structure. E.g. Both cluster '1-1' and cluster '1-2' are subclusters from cluster '1' after first split; while cluster '1-1' and cluster '2-1' come from different cluster after first split.  The second element is a matrix with samples as rows and cluster index as columns, where each entry contains the p-value corresponding to each cell at each split.

This simplified example only contains two clusters, with cluster label either '1' or '2'.

```{r example2, message=FALSE, warning=FALSE}

set.seed(1234)
para1 <- matrix(c(rep(2L, 20), rep(1L, 40)), ncol = 1)
para2 <- matrix(c(rep(2L, 10), rep(1L, 20)), nrow = 1)
dat <- para1 %*% para2
dat[31:60, 1:10] <- dat[31:60, 1:10] + 10L
counts <- map_dbl(dat, ~rpois(1, .x)) %>% 
    matrix(ncol = 30)
storage.mode(counts) <- "integer" 
colnames(counts) <- paste0("c", 1:30)
rownames(counts) <- paste0("g", 1:60)
scppp_obj <- scppp(t(counts))
heatmap(scppp_obj[["data"]])
res <- HclustDepart(scppp_obj, maxSplit = 3)

# cluster results for each cell after each split
head(res[[1]])

# p-value for each cell after each split
head(res[[2]])

```

Please refer to vignette for a more detailed guide, including the application in the real data.

Another option is: keep departure as data representation but apply the Louvain algorithm (implemented in Seurat pipeline) for clustering. 

```{r example3, message=FALSE, warning=FALSE}

res2 <- logit_Louvain_clustering(scppp_obj)
head(res2[[4]])

```

## Issues and bug reports

Please report any issues at ???.

## References


